Finally Tagless, Partially Evaluated

### 2008-08-28 tagless/tagless-final.tex
mostly comments in the notes; while we had figured out what to do about most of Rev2's comments in SF, it has been too long and I have forgotten.
many more updates to the paper - although unfortunately even more notes.

If we evaluate only closed terms that have been type-checked, then
|eval| would never fail. Alas, this soundness is not obvious to the
metalanguage, whose type system we must still appease with the
nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
types above fail to express in the metalanguage that the object program
is well-typed.  This failure necessitates tagging and nonexhaustive
pattern\hyp matching operations that incur a performance penalty in
interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
\citep{taha-tag}.  In short, the universal\hyp type solution is
unsatisfactory because it does not preserve typing.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS, so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.\fi

@@

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types.
\begin{proposition}
If an object term has the object type~$t$, then its
representation in the metalanguage has the type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, 
such as if we forget |app| somewhere above.

@@

@@

Having abstracted our term representation over the interpreter, we are
now ready to present a series of interpreters.  Each interpreter is an
instance of the |Symantics| class in Haskell and a module implementing
the |Symantics| signature in MetaOCaml.

@@

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\ifshort It \else
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it
\fi is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
\ifshort\else
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\fi
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
\ifshort\else
These propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
    The compiler |C| generates tagless code free of pattern matching.
\end{proposition}
This is syntactically clear from the body of |C|, and follows
readily by structural induction and the properties of |.~| and
|.< >.|.

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\input{tagless-final-aplas}
%------------------------------------------------------------------------
% This is the longer version of the section
\else

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

@@

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot
in a large language with more base types and primitive operations.
Incidentally, the code actually contains a more general implementation
mechanism for such features, inspired in part by previous work in
generative linear algebra~\citep{CaretteKiselyov05}.

>>>>

If we evaluate only closed terms that have been type-checked, then
|eval| would never fail. Alas, this soundness is not obvious to the
metalanguage, whose type system we must still appease with the
nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
types above fail to express in the metalanguage that the object program
is well-typed.  This failure necessitates tagging and nonexhaustive
pattern\hyp matching operations that incur a performance penalty in
interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
\citep{taha-tag}.  In short, the universal\hyp type solution is
unsatisfactory because it does not reflect the type of the encoded term.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        even call-by-name (CBN) CPS, so as to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is
available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. 
\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
all examples have also been implemented in Haskell.
\else Except for the fundamental definitions in \S\ref{encoding},
our examples will be shown in (Meta)OCaml, 
even though we have implemented each example equivalently in
both (Meta)OCaml and Haskell%
\footnote{an earlier version of this paper switched liberally between 
both languages, as we believed that our claims were made more obvious
by a judicious choice of metalanguage, but the reviewers disagreed,
pointing out that few people are equally conversant in both.}.
We have chosen to use mainly MetaOCaml because it provides
convenient, typed staging facilities.  We will comment on the significant
differences between the implementations.
\fi

@@

@@

Comparing |Symantics| with Fig.~\ref{fig:object}
shows how to represent \emph{every} typed, closed object term in the
metalanguage. Moreover, the representation preserves types.
\begin{proposition}\label{prop:typing}
If an object term has object type~$t$, then its
representation in the metalanguage has type 
|forall repr.| |Symantics repr => repr |$t$.
\end{proposition}

\begin{proof}
The rules for deriving a type $t$ from a term $T$ of the object language
correspond \emph{bijectively} to the rules encoded in the signature
|Symantics| (when the latter is read as a set of implications, \`{a} la
Curry-Howard).  In other words, each of the rules from Figure~\ref{fig:object}
corresponds bijectively to one line of |Symantics|.

It is worth noting that, as we use HOAS, rules that normally depend on 
context, in other words rules which are themselves an encoding of a function
from the context to a typing relation, (like those for |lam| and |fix|)
correspond to actual metalanguage functions.
\end{proof}
Conversely, the type system of the metalanguage statically checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
object terms, such as if we forget |app| somewhere above.

For simplicity, all examples in the rest of this paper will be in
(Meta)OCaml, but the accompanying code shows a full Haskell implementation
as well.

@@

Now that our term representation is independent of any particular interpreter,
we are now ready to present a series of interpreters.  Each interpreter is an
instance of the |Symantics| class in Haskell and a module implementing
the |Symantics| signature in MetaOCaml.

@@

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
It
\begin{comment}
%% commenting out this Haskell in JFP version as well
\ifshort It \else
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it
\fi 
\enc{comment}
is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If an object term~$e$ encoded (as $e'$) in the metalanguage  has (object)
type~$t$, then evaluating~$e'$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
\ifshort\else
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\fi
\begin{proposition}
  If, for an implementation |I| of |Symantics|, if for all object
  terms $e$ (encoded as $e'$), evaluating $e'$ with |I| always
  terminates, then 
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
  In other words, the (total) relation induced by |I| is a sub-relation
  of the one expressed by the type system of the object language.
  \jacques{Please check that I have this right}
\end{proposition}
\ifshort\else
These propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

\medskip
\caption{The tagless staged interpreter \texttt{C}}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
The environment classifiers here, like the tuple types in \S\ref{ourapproach},
make it a type error to run an open expression.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
    For all object terms $e$ of object type $t$ (encoded as $e'$),
    evaluating $e'$ via the compiler |C| always terminates and
    generates tagless code (i.e. of type $t$ |code|) free of pattern matching.
\end{proposition}
This is syntactically clear from the body of |C|, and follows
readily by structural induction and the properties%
\footnote{which are not easy to establish \citep{MetaML}} of |.~| and
|.< >.|.

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper.
\ifshort
\input{tagless-final-aplas}
%------------------------------------------------------------------------
% This is the longer version of the section
\else

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the type definition for |repr_pe| given
as the start of this section, with |'sv| taking the place of |repr_pe t|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

@@

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot
in a large language with more base types and primitive operations.
Incidentally, the accompanying code actually contains a more general
implementation mechanism for such features, inspired in part by previous work
in generative linear algebra~\citep{CaretteKiselyov05}.
